---
title: "The detection ceiling: what a design could ever have shown"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The detection ceiling: what a design could ever have shown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Tool null or world null

Null results are common in precision psychotherapy, and they are
ambiguous in a specific way. A study that finds no patient-therapist
matching effect has learned one of two very different things: either
the effect is absent in the world, or the design could never have
seen it. Before a null is allowed to mean anything, someone has to
ask what the design was capable of showing.

`design_ceiling()` answers that question retrospectively. Given the
structure of an existing study, it walks an ascending grid of effect
sizes and reports the smallest effect the design detects with a
target power: the design's detection ceiling. Effects below the
ceiling were never observable. A null from such a design is a fact
about the design, not about the world.

## A routine-care structure

Routine settings share a recognizable shape: many therapists, small
caseloads, few measured sessions. Here is a stylized version of that
shape, 60 therapists with 5 patients each and four measured
sessions, diagnosed for a matching effect (the patient-by-therapist
interaction that assignment policies would exploit):

```{r ceiling-routine}
library(tragweite)

routine <- design_ceiling(
  effect       = "matching",
  effect_grid  = seq(0.2, 0.6, by = 0.1),
  n_therapists = 60,
  caseload     = 5,
  n_sessions   = 4,
  icc          = 0.10,
  reps         = 60,
  seed         = 2026,
  quiet        = TRUE
)
routine
```

The grid holds the full diagnosis, with a Monte Carlo standard
error for every rate:

```{r}
routine$grid
```

Two reading rules. First, the ceiling is the design's property, not
the effect's: reporting "no matching effect" from this study means
"no matching effect at least as large as the ceiling". Second, the
grid is computed with common random numbers (the same base seed in
every cell), so the power curve is smooth in the effect size and
differences along the grid are not simulation noise.

## Where the ceiling moves: measurement, not recruitment

The instinctive repair for an underpowered design is more patients.
For nested designs the cheaper lever is usually measurement. Keep
the recruitment identical, 300 patients in the same caseload
structure, and double the measured sessions:

```{r ceiling-repaired}
repaired <- design_ceiling(
  effect       = "matching",
  effect_grid  = seq(0.2, 0.6, by = 0.1),
  n_therapists = 60,
  caseload     = 5,
  n_sessions   = 8,
  icc          = 0.10,
  reps         = 60,
  seed         = 2026,
  quiet        = TRUE
)
repaired
```

```{r, echo = FALSE}
vgl <- data.frame(
  effect   = routine$grid$effect,
  routine  = routine$grid$power,
  repaired = repaired$grid$power
)
knitr::kable(vgl, digits = 2,
             caption = "Detection power, 4 vs. 8 measured sessions (same 300 patients).")
```

Restructuring the same recruitment (fewer therapists with larger
caseloads) moves these curves far less than measurement does; you
can verify that in one line by swapping `n_therapists` and
`caseload` arguments. The pattern is general for this family of
designs: what the analysis starves for is not more people but more
signal per person.

## Honest use

- `reps = 60` keeps this vignette quick; for a diagnosis you intend
  to publish, use `reps = 200` or more and let the reported Monte
  Carlo standard errors say when enough is enough.
- The ceiling is relative to the simulated world: an additive
  product-form matching effect on a standardized outcome, the
  package's transparent default generator. It prices the design's
  structure (nesting, caseloads, sessions, ICC), not every
  possible world.
- `seed` is mandatory by design. A ceiling you cannot reproduce is
  an anecdote.

For prospective planning, the same machinery runs forward:
`tragweite()` for a single design, `design_grid()` for comparing
several under common random numbers, and `min_design()` for the
smallest design that reaches a power target.
