---
title: "The detection ceiling: what a design could ever have shown"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The detection ceiling: what a design could ever have shown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Tool null or world null

Null results are common in precision psychotherapy, and they are
ambiguous in a specific way. A study that finds no patient-therapist
matching effect has learned one of two very different things: either
the effect is absent in the world, or the design could never have
seen it. Before a null is allowed to mean anything, someone has to
ask what the design was capable of showing.

`design_ceiling()` answers that question retrospectively. Given the
structure of an existing study, it walks an ascending grid of effect
sizes and reports the smallest effect the design detects with a
target power: the design's detection ceiling. Effects below the
ceiling fall short of that convention -- they are not unobservable,
and the grid states with what probability each of them was in fact
detected. The further an effect sits below the ceiling, the more a
published null from such a design is a fact about the design rather
than about the world.

## A routine-care structure

Routine settings share a recognizable shape: many therapists, small
caseloads, few measured sessions. Here is a stylized version of that
shape, 60 therapists with 5 patients each and four measured
sessions, diagnosed for a matching effect (the patient-by-therapist
interaction that assignment policies would exploit):

```{r ceiling-routine}
library(tragweite)

routine <- design_ceiling(
  effect       = "matching",
  effect_grid  = seq(0.2, 0.6, by = 0.1),
  n_therapists = 60,
  caseload     = 5,
  n_sessions   = 4,
  icc          = 0.10,
  reps         = 60,
  seed         = 2026,
  quiet        = TRUE
)
routine
```

The grid holds the full diagnosis, with a Monte Carlo standard
error for every rate:

```{r}
routine$grid
```

Two reading rules. First, the ceiling is the design's property, not
the effect's: reporting "no matching effect" from this study means
"no matching effect at least as large as the ceiling". Second, the
grid is computed with common random numbers (the same base seed in
every cell), so the power curve is smooth in the effect size and
differences along the grid are not simulation noise.

## Does more measurement move the ceiling?

The instinctive repair for an underpowered design is more patients.
Measurement is the other candidate lever, and it is usually cheaper.
`design_ceiling()` lets you price it before committing to either.
Keep the recruitment identical, 300 patients in the same caseload
structure, and double the measured sessions:

```{r ceiling-repaired}
repaired <- design_ceiling(
  effect       = "matching",
  effect_grid  = seq(0.2, 0.6, by = 0.1),
  n_therapists = 60,
  caseload     = 5,
  n_sessions   = 8,
  icc          = 0.10,
  reps         = 60,
  seed         = 2026,
  quiet        = TRUE
)
repaired
```

```{r, echo = FALSE}
comparison <- data.frame(
  effect   = routine$grid$effect,
  routine  = routine$grid$power,
  repaired = repaired$grid$power
)
knitr::kable(comparison, digits = 2,
             caption = "Detection power, 4 vs. 8 measured sessions (same 300 patients).")
```

Read that table against the Monte Carlo standard errors in
`routine$grid`, not as a ranking. At `reps = 60` the two columns sit
within about one and a quarter combined standard errors of each
other at every grid point, the cell-by-cell differences change sign,
and **both designs land on the same ceiling of 0.40**. On this evidence,
doubling the measured sessions bought nothing that this run can
distinguish from noise.

That is the honest reading, and it is the point of the exercise
rather than a disappointment: a design lever has to be *priced*
before it is trusted, and here the price came back as "no detectable
effect at this resolution". Whether the lever helps at all is a
question this comparison cannot settle. To settle it, raise `reps`
until the standard errors are smaller than the difference you care
about, and compare the ceilings again; with `reps = 60` a difference
of less than roughly 0.15 in power is simply not resolvable. The
same discipline applies to the recruitment lever: swapping
`n_therapists` and `caseload` produces another comparison that must
be read against its own standard errors before it means anything.

## Honest use

- `reps = 60` keeps this vignette quick; for a diagnosis you intend
  to publish, use `reps = 200` or more and let the reported Monte
  Carlo standard errors say when enough is enough.
- The ceiling is relative to the simulated world: an additive
  product-form matching effect on a standardized outcome, the
  package's transparent default generator. It prices the design's
  structure (nesting, caseloads, sessions, ICC), not every
  possible world.
- `seed` is mandatory by design. A ceiling you cannot reproduce is
  an anecdote.

For prospective planning, the same machinery runs forward:
`tragweite()` for a single design, `design_grid()` for comparing
several under common random numbers, and `min_design()` for the
smallest design that reaches a power target.
